
---
title: "scRNA-seq Analysis for Kameneva Lab PhD Selection"
author: "Adam Adonyi"
date: "April 2025"
output:
  pdf_document:
    fig_width: 10
    fig_height: 6
    fig_caption: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.show = 'hold',
  eval = TRUE
)
```



# SECTION_0: Preparation of environment: loading all essential packages

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(hdf5r)
  library(dplyr)
  library(ggplot2)
  library(gridExtra)
  library(grid)
  library(tidyr)
  library(SingleR)
  library(pheatmap)
  library(patchwork)
  library(SeuratDisk)
  library(DoubletFinder)
  library(scDblFinder)
  library(SingleCellExperiment)
  library(org.Hs.eg.db)  
  library(AnnotationDbi)
  library(GO.db)
  library(celldex)
  library(BiocParallel)
  library(slingshot)
  library(tradeSeq)
  library(SCENIC)
  library(AUCell)
  library(RcisTarget)
  library(conos)
  library(openxlsx)
  library(readxl)
  library(scales)
  library(writexl)
  library(pagoda2)
  library(scater) 
})


```

# SECTION_1: handling RAW.tar file, generating seurat object with all data, save as backup

Download the datasets following the link:
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE147821
Sample name includes the developmental stage in weeks and the internal ID.
week8_001 – this sample is collected from week 8 of development and ID is 001.
Some developmental stages have replicates.

Rendering is unable to handl the pre processing part therefore I had to comment out when I generated the pdf from my Markdown.

Taken together, I opened up the RAW.tar file then loaded all h5 file to generate seurat objects then combined them.
To be able to track the samples I added metadata furthermore I calculated and assigned mitocondrial content.

```{r}
# processing RAW.tar file to extracted format

#file.exists("~/Downloads/GSE147821_RAW.tar")
#untar("Downloads/GSE147821_RAW.tar", exdir = "Downloads/GSE147821_extracted")
```

```{r}
# # Generate Seurat object with metadata, combine them into one object and calculate mitochondrial content (mt) as first step of QC

# # folder_path <- "/Users/adamadonyi/Downloads/GSE147821_extracted"

# # Define GSM to week_id mapping
# sample_names <- c(
#   "GSM4446535" = "week8_001",
#   "GSM4446536" = "week9_063",
#   "GSM4446537" = "week6_088",
#   "GSM4446538" = "week14_123",
#   "GSM4446539" = "week12_124",
#   "GSM4446540" = "week8_125",
#   "GSM4446541" = "week9_005",
#   "GSM4446542" = "week11_006",
#   "GSM4446543" = "week9_007",
#   "GSM4734601" = "week8_016",
#   "GSM4734602" = "week9_031_paraganglia",
#   "GSM4734603" = "week12_035",
#   "GSM4734604" = "week12_036_extraadrenal"
# )

# # List all H5 files 
# file_list <- list.files(path = folder_path, pattern = "*.h5$", full.names = TRUE)

# # Initialize a list 
# seurat_list <- list()

# # Process each file and create Seurat objects
# for (file_path in file_list) {
#   # Extract GSM from the file path
#   gsm <- gsub(".*/(GSM[0-9]+)_.*", "\\1", file_path)
  
#   # Get the sample name
#   sample_name <- sample_names[gsm]
  
#   # Extract week and ID
#   week <- as.numeric(gsub("week([0-9]+)_.*", "\\1", sample_name))
#   id <- gsub("week[0-9]+_([0-9]+.*)", "\\1", sample_name)
  
#   # Print progress to be able to track process
#   cat("Processing:", gsm, "as", sample_name, "- Week:", week, "ID:", id, "\n")
  
#   # Read the 10X h5 file --> Seurat object
#   tryCatch({
#     # Read data
#     data <- Read10X_h5(file_path)
    
#     # Create initial Seurat object
#     seurat_obj <- CreateSeuratObject(
#       counts = data,
#       project = sample_name,
#       min.cells = 3,
#       min.features = 200
#     )
    
#     # Add metadata !
#     # First create a data frame with metadata for each cell
#     cell_ids <- colnames(seurat_obj)
#     metadata_df <- data.frame(
#       gsm = rep(gsm, length(cell_ids)),
#       sample = rep(sample_name, length(cell_ids)),
#       week = rep(week, length(cell_ids)),
#       id = rep(id, length(cell_ids)),
#       row.names = cell_ids
#     )
    
#     # Add metadata to Seurat object
#     seurat_obj <- AddMetaData(seurat_obj, metadata = metadata_df)
    
#     # Add to list (add comment to be able to track the loading)
#     seurat_list[[gsm]] <- seurat_obj
    
#     cat("Successfully processed", gsm, "\n")
#   }, error = function(e) {
#     cat("Error processing", gsm, ":", e$message, "\n")
#   })
# }

# # Merge all Seurat objects into one (only if there are objects in the list, > for that)
# if (length(seurat_list) > 0) {
#   # Merge all objects
#   cat("Merging", length(seurat_list), "Seurat objects...\n")
  
#   if (length(seurat_list) == 1) {
#     combined_seurat <- seurat_list[[1]]
#   } else {
#     combined_seurat <- merge(
#       x = seurat_list[[1]],
#       y = seurat_list[-1],
#       add.cell.ids = names(seurat_list),
#       project = "GSE147821"
#     )
#   }
  
#   # Calculate percentage of mitochondrial genes as first step of QC
#   combined_seurat[["percent.mt"]] <- PercentageFeatureSet(combined_seurat, pattern = "^MT-")
  
#   # Print summary of the combined object
#   cat("\nCombined Seurat object:\n")
#   print(combined_seurat)
  
#   # Check metadata columns
#   cat("\nMetadata columns:\n")
#   print(colnames(combined_seurat@meta.data))
  
#   # Show a sample of the metadata
#   cat("\nSample of metadata:\n")
#   print(head(combined_seurat@meta.data))
# } else {
#   cat("No Seurat objects were created. Check for errors above.\n")
# }

# # Remove the "sample" column from metadata since it is duplicated
# combined_seurat$sample <- NULL

```

```{r}
# Inspection of object 
#head(combined_seurat@meta.data)
```


```{r}
# All data should be loaded (13)
#unique(combined_seurat@meta.data$orig.ident)

```


```{r}
# Save the Seurat object as check point
#saveRDS(combined_seurat, file = "/Users/adamadonyi/Downloads/GSE147821_combined_seurat.rds")
```



# SECTION_2: QC and cell cycle correction
Perform data normalization (consider cell cycle correction) and quality control,
data cleaning.

First I visualized each sample to inspect the nCount_RNA, nFeature_RNA and mitocondrial content



```{r}
# Load previously saved data (combined objects with the calculated mt content)
combined_seurat <- readRDS("/Users/adamadonyi/Downloads/GSE147821_combined_seurat.rds")

```

```{r}
# define specific order of the samples
custom_order <- unique(combined_seurat@meta.data[order(combined_seurat@meta.data$week), "orig.ident"])

```

```{r}
# visualization of nCount_RNA
VlnPlot(combined_seurat, features = c("nCount_RNA"), 
        group.by = "orig.ident", 
        pt.size = 0) + 
  scale_x_discrete(limits = custom_order) +  # Reorder X-axis
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
# visualization of nFeature_RNA
VlnPlot(combined_seurat, features = c("nFeature_RNA"), 
        group.by = "orig.ident", 
        pt.size = 0) + 
  scale_x_discrete(limits = custom_order) +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# visualization of percent.mt
VlnPlot(combined_seurat, features = c("percent.mt"), 
        group.by = "orig.ident", 
        pt.size = 0) + 
  scale_x_discrete(limits = custom_order) +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
To be able to define thresholds for QC sub setting, I implement three time median absolute deviation on nCount_RNA, nFeature_RNA, and percent.mt. The calculated thresholds helped to define the parameters for sub setting.

```{r}
# Calculate thresholds using median absolute deviation (MAD)
max.count.thr <- median(combined_seurat$nCount_RNA) + 3*mad(combined_seurat$nCount_RNA)
min.count.thr <- median(combined_seurat$nCount_RNA) - 3*mad(combined_seurat$nCount_RNA)

max.feature.thr <- median(combined_seurat$nFeature_RNA) + 3*mad(combined_seurat$nFeature_RNA)
min.feature.thr <- median(combined_seurat$nFeature_RNA) - 3*mad(combined_seurat$nFeature_RNA)

max.mito.thr <- median(combined_seurat$percent.mt) + 3*mad(combined_seurat$percent.mt)
min.mito.thr <- median(combined_seurat$percent.mt) - 3*mad(combined_seurat$percent.mt)
```

After calculation, I visualized the thresholds with their corresponding data category.

```{r}
# Create feature count plot
g_feature <- ggplot(as.data.frame(combined_seurat@meta.data))
g_feature <- g_feature + geom_histogram(aes(x=nFeature_RNA, y= ..density..),bins=200,fill = "grey")
g_feature <- g_feature + geom_density(aes(x=nFeature_RNA)) 
g_feature <- g_feature + 
  geom_vline(xintercept = min.feature.thr, color="red") +
  geom_vline(xintercept = max.feature.thr, color="blue") +
  ggtitle("Gene Features") +
  theme_minimal()

# Create count plot
g_count <- ggplot(as.data.frame(combined_seurat@meta.data))
g_count <- g_count + geom_histogram(aes(x=nCount_RNA, y= ..density..),bins=200,fill = "grey")
g_count <- g_count + geom_density(aes(x=nCount_RNA)) 
g_count <- g_count + 
  geom_vline(xintercept = min.count.thr, color="red") + 
  geom_vline(xintercept = max.count.thr, color="blue") +
  ggtitle("RNA Counts") +
  theme_minimal()

# Create mitochondrial percentage plot
g_mito <- ggplot(as.data.frame(combined_seurat@meta.data))
g_mito <- g_mito + geom_histogram(aes(x=percent.mt, y= ..density..),bins=200,fill = "grey")
g_mito <- g_mito + geom_density(aes(x=percent.mt)) 
g_mito <- g_mito + 
  geom_vline(xintercept = min.mito.thr, color="red") + 
  geom_vline(xintercept = max.mito.thr, color="blue") +
  ggtitle("Mitochondrial %") +
  theme_minimal()
```

```{r}
# display them in a grid arrangement
grid_plot <- grid.arrange(g_feature, g_count, g_mito, ncol = 3)
```

```{r}
# Print threshold values for reference
cat("Count thresholds:", min.count.thr, "-", max.count.thr, "\n") #1000 - 18054 -->but 15000?
cat("Feature thresholds:", min.feature.thr, "-", max.feature.thr, "\n") #200-3700
cat("Mitochondrial % thresholds:", min.mito.thr, "-", max.mito.thr, "\n") # 0-23.10


```

Count thresholds: -10864.8 & 18054.8 --> used from 1000 to 15 000 (I thought 18 000 would increase the chance to introduce too much noise to the data which can cause problems later. Furthermore, I had to keep in mind the lack of computational power since I run on a local pc. I did not do any comparison because of the lack of time. It would be nice to investigate the differences)

Feature thresholds: I used the upper calculated threshold. Lower bond for 200 is from the standard Seurat pipeline recommendation. Upper bond is 3700.

Mitochondrial percentage thresholds: According to the visual-given gut feeling, 25% would be the best cut for me, and MAD supported the same range therefore went for the calculated 23.1.

I used those numbers for sub setting my data set.

```{r}
# Subset with the mentioned thresholds and going into section 3
combined_seurat <- subset(combined_seurat, subset = nFeature_RNA > 200 & nFeature_RNA < 3700 & percent.mt < 23.1 & nCount_RNA > 1000 & nCount_RNA < 15000)

```


SECTION_2.2: Cell cycle correction and final steps of standard scRNAseq pipeline

Seurat build in gene collection supports us to implement the cell cycle scoring. I used to assign potential cell cycle status to each cell. Visualized full data set before and after the regression. (Original documentation recommend PC on the suppressed genes only - which did not work for me. As an alternative solution I visualized the PC1 & PC2 using the full data set which showed alteration therefore the regression was a proper step). 

```{r}
# cell cycle correction

combined_seurat <- NormalizeData(combined_seurat)
DefaultAssay(combined_seurat) <- "RNA"

# Join the layers  (as suggested by the error message!)
combined_seurat <- JoinLayers(object = combined_seurat)

# Assigning genes from Seurat package for cell cycle scoring
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

# Run the Scoring 
combined_seurat <- CellCycleScoring(combined_seurat, 
                                    s.features = s.genes, 
                                    g2m.features = g2m.genes)

# Visualize the cell cycle effect before correction 
combined_seurat <- FindVariableFeatures(combined_seurat)
combined_seurat <- ScaleData(combined_seurat)
combined_seurat <- RunPCA(combined_seurat) # running the PCA only on the s.genes and g2m.genes does not work therefore I use full data
DimPlot(combined_seurat, reduction = "pca", group.by = "Phase")
```



```{r}
# Regress out cell cycle effects during scaling
combined_seurat <- ScaleData(combined_seurat, 
                             vars.to.regress = c("S.Score", "G2M.Score"))

# Run PCA again and check if cell cycle effect is reduced
combined_seurat <- RunPCA(combined_seurat) # running the PCA only on the s.genes and g2m.genes does not work 
DimPlot(combined_seurat, reduction = "pca", group.by = "Phase")


```
DoubletFinder DID NOT WORK - I REACHED THE MAX 16GB RAM therefore I did not use the package here as next step - HPC could provide enough computational memory for the data set. Other option would be parallel computing but I did not spend time on the optimization. Doublet removal will be implemented when I subset the data to investigate only the SCP, chromaffin and sympathoblast populations.

```{r}

# get the latest version
#if (!requireNamespace("remotes", quietly = TRUE)) {
#  install.packages("remotes")
#}
#remotes::install_github("chris-mcginnis-ucsf/DoubletFinder")

#library(DoubletFinder)

# Run after normalization and PCA but before clustering!!!!
# For the current version, use these functions:
#sweep.res <- paramSweep(combined_seurat, PCs = 1:15, sct = FALSE)
#sweep.stats <- summarizeSweep(sweep.res, GT = FALSE)
#bcmvn <- find.pK(sweep.stats)
#optimal_pk <- bcmvn$pK[which.max(bcmvn$BCmetric)]

# Run DoubletFinder 
#combined_seurat <- doubletFinder(combined_seurat, 
#                                 PCs = 1:15, 
#                                 pN = 0.05,         # adjusted from 0.25 overloaded the RAM - DO NOT WORK-----
#                                 pK = optimal_pk, 
#                                 nExp = round(0.075*ncol(combined_seurat)))


```

ElbowPlot shows the PCs with standard deviation to pick the optimal dimensionality. 

```{r}
# Elbow to determine the PCs for the next steps
ElbowPlot(combined_seurat)

# tested with 15 but went for 10 because of computational manner.
```

First I ran resolution=0.5 then 0.3 but ended up using 0.2 to get the proper depth. 0.5 gives more than 23 clusters which could lead to a lot of confusion in case of annotation.

```{r}
# Further steps of standard scRNAseq pipeline -  dims 10, cluster resolution 0.2
combined_seurat <- FindNeighbors(combined_seurat, dims = 1:10)
combined_seurat <- FindClusters(combined_seurat, resolution = 0.2) # adjusted resolution from default 0.5 (tested, 0.3, but used 0.2)
combined_seurat <- RunUMAP(combined_seurat, dims = 1:10)

```

Visualization section after finishing the standard scRNAseq pipeline comparing orig.idents, time course (weeks), and identified clusters using UMAP.

```{r}
# Visual based on the origi.ident
DimPlot(combined_seurat, reduction = "umap", group.by = "orig.ident")

```

```{r}
# Visual based on the time course
DimPlot(combined_seurat, reduction = "umap", group.by  = "week")
```


```{r}
# Visual based on the identified clusters
DimPlot(combined_seurat, reduction = "umap")

```
Overall, 15 Cluster identified (resolution 0.5 resulted more than 23 clusters which provide a deeper insight to the data but for my purpose 15 cluster should be enough). Considering the biological composition based on the given papers, the 15 clusters seemed reasonable for further analysis. 

To further investigate clusters and their signature markers, I was run FindAllMarkers comparing all clusters with all then showed the top 5 genes. This graph is going to be shown in the section 4 following the structure of the instruction from the given mail.

Considering the restricted access to computational power I was focusing on only the positively regulated genes. 
Optimal speed up solution for further investigation would be defining min.pct (0.1-0.3) which only test genes that are detected in a minimum fraction of min.pct cells in either of the two populations. (default is 0.01)



As end of the section, I saved the object as rds in case I have to restart from that step.

```{r}
# End of section 3, saved rds as a backout
#saveRDS(combined_seurat, file = "/Users/adamadonyi/Downloads/GSE147821_combined_seurat_clusteres.rds")

# remove previous object and free memory
rm(combined_seurat)
gc()
```

# SECTION_3: Generate the UMAP and perform cluster annotation.

```{r}
combined_seurat <- readRDS("/Users/adamadonyi/Downloads/GSE147821_combined_seurat_clusteres.rds")
```

```{r}
# Visualization of all clusters (not annotated yet)
DimPlot(combined_seurat, reduction = "umap")

```

Standard annotation method is using already established packages with reference data sets. The most heavily used and documented the singleR which is able to call different (human, mouse) references. The best method to handle human immune cells therefore I am familiar with the package but here HumanPrimaryCellAtlasData (which is the most general and broad library) failed to capture most of the cell types. After manual annotation I can confirm that singleR guessed correctly only 3 cell types. Leave the code here for reviewing in case I made some mistake, or adjustment is possible.

```{r}
#ref <- celldex::HumanPrimaryCellAtlasData()
#colData(ref)

#result <- SingleR(test = as.SingleCellExperiment(combined_seurat), ref = ref, labels = ref$label.main)

#combined_seurat$singlr_labels <- result$labels

#View(combined_seurat@meta.data)

#saveRDS(combined_seurat, file = "/Users/adamadonyi/Downloads/GSE147821_combined_seurat_clusteres_singler.rds")
#combined_seurat <- readRDS("/Users/adamadonyi/Downloads/GSE147821_combined_seurat_clusteres_singler.rds")
#DimPlot(combined_seurat, reduction = "umap", group.by = 'singlr_labels', label = TRUE, repel = TRUE)

```

I approached the task manually based on the given reference paper, I was able to annotate all clusters, however, I think, my resolution was too high and sometimes I had to assigned 2 identified clusters to one cell type but I did it only after confirming via visualizing the cell type specific genes and clusters simultaneously. 



```{r}
Idents(combined_seurat) <- "seurat_clusters"
```



Cluster 0 with Endothelium markers

```{r umap_combined_plots_0, fig.width=12, fig.height=5}


# UMAP with highlighted cluster 0
umap_highlight_0 <- DimPlot(
  combined_seurat,
  reduction = "umap",
  cells.highlight = WhichCells(combined_seurat, idents = "0"),
  cols.highlight = "black",
  cols = "lightgrey"
) + ggtitle("UMAP: Highlighting Cluster 0")

# Feature plots for two marker genes
feature_plot_0 <- FeaturePlot(
  combined_seurat,
  reduction = "umap",
  features = c("PECAM1", "KDR")
)

# Combine them and display
umap_highlight_0 + feature_plot_0

```

Cluster 1 with Adrenal gland cortex marker

```{r umap_combined_plots_1, fig.width=12, fig.height=5}


# UMAP with highlighted cluster 
umap_highlight_1 <- DimPlot(
  combined_seurat,
  reduction = "umap",
  cells.highlight = WhichCells(combined_seurat, idents = "1"),
  cols.highlight = "black",
  cols = "lightgrey"
) + ggtitle("UMAP: Highlighting Cluster 1")

# Feature plots for two marker genes
feature_plot_1 <- FeaturePlot(
  combined_seurat,
  reduction = "umap",
  features = c("NR5A1")
)

# Combine them and display
umap_highlight_1 + feature_plot_1

```

Cluster 2 with Subepicardial and abdominal mesenchyme marker

```{r umap_combined_plots_2, fig.width=12, fig.height=5}


# UMAP with highlighted cluster 
umap_highlight_2 <- DimPlot(
  combined_seurat,
  reduction = "umap",
  cells.highlight = WhichCells(combined_seurat, idents = "2"),
  cols.highlight = "black",
  cols = "lightgrey"
) + ggtitle("UMAP: Highlighting Cluster 2")

# Feature plots for two marker genes
feature_plot_2 <- FeaturePlot(
  combined_seurat,
  reduction = "umap",
  features = c("PRRX1")
)

# Combine them and display
umap_highlight_2 + feature_plot_2

```
Cluster 3 with the Erythroid cells marker
```{r umap_combined_plots_3, fig.width=12, fig.height=5}


# UMAP with highlighted cluster 0
umap_highlight_3 <- DimPlot(
  combined_seurat,
  reduction = "umap",
  cells.highlight = WhichCells(combined_seurat, idents = "3"),
  cols.highlight = "black",
  cols = "lightgrey"
) + ggtitle("UMAP: Highlighting Cluster 3")

# Feature plots for two marker genes
feature_plot_3 <- FeaturePlot(
  combined_seurat,
  reduction = "umap",
  features = c("HBA2", "HBB")
)

# Combine them and display
umap_highlight_3 + feature_plot_3

```
Cluster 4 with the Erythroid cells marker

```{r umap_combined_plots_4, fig.width=12, fig.height=5}


# UMAP with highlighted cluster 0
umap_highlight_4 <- DimPlot(
  combined_seurat,
  reduction = "umap",
  cells.highlight = WhichCells(combined_seurat, idents = "4"),
  cols.highlight = "black",
  cols = "lightgrey"
) + ggtitle("UMAP: Highlighting Cluster 4")

# Feature plots for two marker genes
feature_plot_4 <- FeaturePlot(
  combined_seurat,
  reduction = "umap",
  features = c("HBA2", "HBB")
)

# Combine them and display
umap_highlight_4 + feature_plot_4

```

Cluster 5 with Sympathoblasts marker

```{r umap_combined_plots_5, fig.width=12, fig.height=5}


# UMAP with highlighted cluster 0
umap_highlight_5 <- DimPlot(
  combined_seurat,
  reduction = "umap",
  cells.highlight = WhichCells(combined_seurat, idents = "5"),
  cols.highlight = "black",
  cols = "lightgrey"
) + ggtitle("UMAP: Highlighting Cluster 5")

# Feature plots for two marker genes
feature_plot_5 <- FeaturePlot(
  combined_seurat,
  reduction = "umap",
  features = c("STMN2")
)

# Combine them and display
umap_highlight_5 + feature_plot_5

```
Cluster 6 with the Immune cell markers

```{r umap_combined_plots_6, fig.width=12, fig.height=5}


# UMAP with highlighted cluster 0
umap_highlight_6 <- DimPlot(
  combined_seurat,
  reduction = "umap",
  cells.highlight = WhichCells(combined_seurat, idents = "6"),
  cols.highlight = "black",
  cols = "lightgrey"
) + ggtitle("UMAP: Highlighting Cluster 6")

# Feature plots for two marker genes
feature_plot_6 <- FeaturePlot(
  combined_seurat,
  reduction = "umap",
  features = c("FCGR1A", "CD163")
)

# Combine them and display
umap_highlight_6 + feature_plot_6

```
Cluster 7 with the Liver markers
```{r umap_combined_plots_7, fig.width=12, fig.height=5}


# UMAP with highlighted cluster 0
umap_highlight_7 <- DimPlot(
  combined_seurat,
  reduction = "umap",
  cells.highlight = WhichCells(combined_seurat, idents = "7"),
  cols.highlight = "black",
  cols = "lightgrey"
) + ggtitle("UMAP: Highlighting Cluster 7")

# Feature plots for two marker genes
feature_plot_7 <- FeaturePlot(
  combined_seurat,
  reduction = "umap",
  features = c("HNF4A", "AHSG"),
)

# Combine them and display
umap_highlight_7 + feature_plot_7

```
Cluster 8 with "Chromaffin cell markers 
```{r umap_combined_plots_8, fig.width=12, fig.height=5}


# UMAP with highlighted cluster 0
umap_highlight_8 <- DimPlot(
  combined_seurat,
  reduction = "umap",
  cells.highlight = WhichCells(combined_seurat, idents = "8"),
  cols.highlight = "black",
  cols = "lightgrey"
) + ggtitle("UMAP: Highlighting Cluster 8")

# Feature plots for two marker genes
feature_plot_8 <- FeaturePlot(
  combined_seurat,
  reduction = "umap",
  features = c("ELAVL3", "ELAVL4", "PHOX2B", "TH")
)

# Combine them and display
umap_highlight_8 + feature_plot_8
```
Cluster 9 with SCPs markers
```{r umap_combined_plots_9, fig.width=12, fig.height=5}

# UMAP with highlighted cluster 
umap_highlight_9 <- DimPlot(
  combined_seurat,
  reduction = "umap",
  cells.highlight = WhichCells(combined_seurat, idents = "9"),
  cols.highlight = "black",
  cols = "lightgrey"
) + ggtitle("UMAP: Highlighting Cluster 9")

# Feature plots for two marker genes
feature_plot_9 <- FeaturePlot(
  combined_seurat,
  reduction = "umap",
  features = c("SOX10", "PLP1", "FOXD3")
)

# Combine them and display
umap_highlight_9 + feature_plot_9


```

Cluster 10 with Kidney marker 
```{r umap_combined_plots_10, fig.width=12, fig.height=5}


# UMAP with highlighted cluster 0
umap_highlight_10 <- DimPlot(
  combined_seurat,
  reduction = "umap",
  cells.highlight = WhichCells(combined_seurat, idents = "10"),
  cols.highlight = "black",
  cols = "lightgrey"
) + ggtitle("UMAP: Highlighting Cluster 10")

# Feature plots for two marker genes
feature_plot_10 <- FeaturePlot(
  combined_seurat,
  reduction = "umap",
  features = c("PAX2")
)

# Combine them and display
umap_highlight_10 + feature_plot_10

```
Cluster 11 with Intermediate mesoderm
```{r umap_combined_plots_11, fig.width=12, fig.height=5}


# UMAP with highlighted cluster 0
umap_highlight_11 <- DimPlot(
  combined_seurat,
  reduction = "umap",
  cells.highlight = WhichCells(combined_seurat, idents = "11"),
  cols.highlight = "black",
  cols = "lightgrey"
) + ggtitle("UMAP: Highlighting Cluster 11")

# Feature plots for two marker genes
feature_plot_11 <- FeaturePlot(
  combined_seurat,
  reduction = "umap",
  features = c("GATA4", "HAND2")
)

# Combine them and display
umap_highlight_11 + feature_plot_11

```
Cluster 12 with Erythroid cell markers
```{r umap_combined_plots_12, fig.width=12, fig.height=5}


# UMAP with highlighted cluster 0
umap_highlight_12 <- DimPlot(
  combined_seurat,
  reduction = "umap",
  cells.highlight = WhichCells(combined_seurat, idents = "12"),
  cols.highlight = "black",
  cols = "lightgrey"
) + ggtitle("UMAP: Highlighting Cluster 12")

# Feature plots for two marker genes
feature_plot_12 <- FeaturePlot(
  combined_seurat,
  reduction = "umap",
  features = c("HBA2", "HBB")
)

# Combine them and display
umap_highlight_12 + feature_plot_12

```
Cluster 13 with Liver markers c("HNF4A", "AHSG"),
```{r umap_combined_plots_13, fig.width=12, fig.height=5}


# UMAP with highlighted cluster 0
umap_highlight_13 <- DimPlot(
  combined_seurat,
  reduction = "umap",
  cells.highlight = WhichCells(combined_seurat, idents = "13"),
  cols.highlight = "black",
  cols = "lightgrey"
) + ggtitle("UMAP: Highlighting Cluster 13")

# Feature plots for two marker genes
feature_plot_13 <- FeaturePlot(
  combined_seurat,
  reduction = "umap",
  features = c("HNF4A", "AHSG")
)

# Combine them and display
umap_highlight_13 + feature_plot_13

```
Cluster 14 with Immune cells markers
```{r umap_combined_plots_14, fig.width=12, fig.height=5}


# UMAP 
umap_highlight_14 <- DimPlot(
  combined_seurat,
  reduction = "umap",
  cells.highlight = WhichCells(combined_seurat, idents = "14"),
  cols.highlight = "black",
  cols = "lightgrey"
) + ggtitle("UMAP: Highlighting Cluster 14")

# Feature plots for two marker genes
feature_plot_14 <- FeaturePlot(
  combined_seurat,
  reduction = "umap",
  features = c("FCGR1A", "CD163")
)

# Combine them and display
umap_highlight_14 + feature_plot_14

```
Cluster 15 with Liver 
```{r umap_combined_plots_15, fig.width=12, fig.height=5}


# UMAP with highlighted cluster 0
umap_highlight_15 <- DimPlot(
  combined_seurat,
  reduction = "umap",
  cells.highlight = WhichCells(combined_seurat, idents = "15"),
  cols.highlight = "black",
  cols = "lightgrey"
) + ggtitle("UMAP: Highlighting Cluster 15")

# Feature plots for two marker genes
feature_plot_15 <- FeaturePlot(
  combined_seurat,
  reduction = "umap",
  features = c("HNF4A", "AHSG")
)

# Combine them and display
umap_highlight_15 + feature_plot_15

```

Assign the identified cell types to the clusters and visualize.

```{r}

# Create a named vector for mapping cluster numbers to labels
cluster_map <- c(
  "0" = "Endothelium",
  "1" = "Adrenal gland cortex",
  "2" = "Subepicardial and abdominal mesenchyme",
  "3" = "Erythroid cells",
  "4" = "Erythroid cells",
  "5" = "Sympathoblasts",
  "6" = "Immune cells",
  "7" = "Liver",
  "8" = "Chromaffin cells",
  "9" = "SCPs",
  "10" = "Kindey",
  "11" = "Intermediate mesoderm",
  "12" = "Erythroid cells",
  "13" = "Liver",
  "14" = "Immune cells",
  "15" = "Liver"
)

# Add the cluster_label column to the Seurat object using the mapping
combined_seurat@meta.data$cluster_label <- cluster_map[combined_seurat@meta.data$seurat_clusters]

Idents(combined_seurat) <- "cluster_label"
DimPlot(combined_seurat, reduction = "umap", 
        group.by = "cluster_label",  # Use your custom labels
        label = TRUE, 
        repel = TRUE) + 
  NoLegend()

```



# SECTION_4: Visualization of differently expressed top 5 genes for each cluster


Running FindAllMarkers provides the differently expressed genes comparing all clusters with all. It is commented out because I was not able to render my markdown to pdf keeping that line of code. I overloaded my RAM. But when I did it manually, during generating the notebook, I saved in xlsx format so I can re-load and use it for visualization. As I mentioned, I was focusing on only the positive ones. 

```{r}
#clusters.markers <- FindAllMarkers(combined_seurat, only.pos = TRUE)
```



```{r}
# generate Excel to be able to use later
#write_xlsx(clusters.markers, path = "clusters_markers.xlsx")
clusters.markers <- read_excel("/Users/adamadonyi/Downloads/clusters_markers.xlsx")

```


```{r}
colnames(clusters.markers)
```


Visualize the top 5 differently expressed genes of each cluster as DotPlot

```{r expressed genes_plot, fig.width=12, fig.height=15}

# Collect top 5 differently expressed genes for all clusters
top_5_no_min <- clusters.markers %>% 
  group_by(cluster) %>% 
  slice_max(n = 5, order_by = avg_log2FC)



DotPlot(combined_seurat, 
        features = top_5_no_min$gene, 
        group.by = "cluster_label") + 
  coord_flip() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.5, size = 8), 
    axis.text.y = element_text(hjust = 1, vjust = 0.5, size = 8)  
  ) + 
  labs(title = "Top Marker Genes by Cluster",
       x = "Genes",
       y = "Clusters")



```

# SECTION_5: Sub-select adrenal medulla clusters (Schwann cell precursors (SCPs), Chromaffin cells, Sympathoblasts) and re-cluster them to improve the resolution of transitions. Note that NOT all datasets may be included in the re-clustering.If you choose to do so, explain why that may be necessary.

Saved the version in case I have to restart from that point.

```{r}
#saveRDS(combined_seurat, file = "/Users/adamadonyi/Downloads/GSE147821_combined_seurat_clusteres_annotated.rds")
```

```{r clear_environment}
# Remove all objects from the environment
rm(list = ls())



# Optionally, you can force garbage collection
gc()
```

```{r}
# Load last object for further work
combined_seurat <- readRDS("/Users/adamadonyi/Downloads/GSE147821_combined_seurat_clusteres_annotated.rds")

```

Before isolation of adrenal medulla cluster some steps are necessary to make sure we have a clean, balanced data set.
1, remove doublets (scDblFinder)
2, remove cell cycle related genes (similarly to the publication from 2021 Kameneva, https://doi.org/10.1038/s41588-021-00818-x, the similar approach considered to introduce the removal of certain cell cycle related genes. This has been done based on the "GO:0000278", "GO:0006260", "GO:0007049").
3, remove cells with STAR genes expressed - which is cortex marker.


```{r}
# process is made on full dataset
# Convert Seurat to SingleCellExperiment
sce <- as.SingleCellExperiment(combined_seurat)

# Run scDblFinder (with batch information))
sce <- scDblFinder(sce, 
                   samples = sce$orig.ident,  # This tells the function about your samples
                   dbr = NULL,  # Let the function estimate doublet rates per sample
                   dims = 30)  # Return all scores and info

# Add results back to Seurat object
combined_seurat$scDblFinder_score <- sce$scDblFinder.score
combined_seurat$scDblFinder_class <- sce$scDblFinder.class
combined_seurat$scDblFinder_keep <- sce$scDblFinder.class == "singlet"

DimPlot(combined_seurat, reduction = "umap", group.by = 'scDblFinder_keep', label = FALSE, repel = TRUE) 


```

```{r}
# distribution of doublets and siglets along the full dataset
doublet_table <- table(combined_seurat$scDblFinder_class)
print(doublet_table)

```

singlet doublet 
  67728    4245 
Which is a reasonable ratio since with 10X platform we can expect doublets between 1-10% depending on the samples.

Computationally impossible to process the cell-cycle-related-genes removal step on the full data set therefore I isolate first the populations of our interest and continue with them.

```{r}
# isolation of "Sympathoblasts", "Chromaffin cells" and "SCPs" 
# First, check the identities
table(Idents(combined_seurat))
colnames(combined_seurat@meta.data)
Idents(combined_seurat) <- "cluster_label"

# Subset the object
sub_seurat <- subset(combined_seurat, idents = c("Sympathoblasts", "Chromaffin cells", "SCPs"))
# Inspect the content of sub setted data
doublet_table <- table(sub_seurat$scDblFinder_class)
print(doublet_table)
table(Idents(sub_seurat))
```
The isolated populations consist of decent amount of cells for further analysis and the singlet doublet ratio is also good. 

As next step, I used the reference cell cycle related GO terms and their genes to remove from my data set.

```{r}
# Get genes associated with cell cycle GO terms
cell_cycle_go_terms <- c("GO:0000278", "GO:0006260", "GO:0007049")
cell_cycle_genes <- unique(unlist(lapply(cell_cycle_go_terms, function(go_term) {
  genes <- AnnotationDbi::select(org.Hs.eg.db, 
                                 keys = go_term,
                                 columns = c("SYMBOL"), 
                                 keytype = "GOALL")$SYMBOL
  return(genes)
})))

# Print how many cell cycle genes were found
print(paste0("Number of cell cycle genes found: ", length(cell_cycle_genes)))
```

1849 cell cycle related genes are in total from the 3 GOs.

```{r}
# Get the current genes in your Seurat object
current_genes <- rownames(sub_seurat)

# Find overlap between your datasets and cell cycle genes
cc_genes_in_data <- intersect(current_genes, cell_cycle_genes)
print(paste0("Number of cell cycle genes in your dataset: ", length(cc_genes_in_data)))

# Create a new Seurat object without cell cycle genes!
genes_to_keep <- setdiff(current_genes, cc_genes_in_data)
sub_seurat_no_cc <- subset(sub_seurat, features = genes_to_keep)

# Compare the number of genes before and after
print(paste0("Genes before filtering: ", nrow(sub_seurat)))
print(paste0("Genes after filtering: ", nrow(sub_seurat_no_cc)))
```

1700 genes have been found in the data set and removed. Confirmed by the before after gene number.

According to the reference paper, genes with high correlation with "MKI67", "TOP2A", "TUBB4B" (based on pearson correlation) also nice to be removed. I had total 4 related genes with those, but these genes were removed after the previous section anyway. (method/code not shown here)

Additionally, focusing on adrenal medulla cluster, removing the STAR gene "contaminated" events is also beneficial since STAR gene usually expressed in cortex.

After sub setting the STAR positive cells the isolated clusters are free from STAR gene.

```{r}
# Direct approach using subset
sub_seurat_no_cc <- subset(sub_seurat_no_cc, subset = STAR == 0)
FeaturePlot(sub_seurat_no_cc, reduction = "umap", features = c("STAR"))
```
From that point the isolated, cleaned sub-clusters is in the focus. The next step is to implement the standard scRNAseq pipeline on the isolated data set.

```{r}
sub_seurat_no_cc <- NormalizeData(sub_seurat_no_cc, normalization.method = "LogNormalize", scale.factor = 10000)

# 2. Find variable features
sub_seurat_no_cc <- FindVariableFeatures(sub_seurat_no_cc, selection.method = "vst", nfeatures = 2000)

# 3. Scale the data
sub_seurat_no_cc <- ScaleData(sub_seurat_no_cc)

# 4. Run PCA - as per paper, they used 30 components
sub_seurat_no_cc <- RunPCA(sub_seurat_no_cc)

# 5. Determine significant PCs --> use the 3
ElbowPlot(sub_seurat_no_cc)

# important step - tested without 
sub_seurat_no_cc <- FindNeighbors(sub_seurat_no_cc, dims = 1:3)
sub_seurat_no_cc <- FindClusters(sub_seurat_no_cc, resolution = 0.1)

# 6 Run UMAP
sub_seurat_no_cc <- RunUMAP(sub_seurat_no_cc, dims = 1:3)


```


Beside of removing the doublets, I decided to remove the extra adrenal labelled sample since we focus on the non-extra-adrenal cells, furthermore continued the analysis without the paraganglia labelled sample because it is not part of the sympatho-adrenal section and could mask important genes.

```{r}
# Subset to exclude unwanted samples (mentioned  - 2) and remove doublets
sub_seurat_no_cc <- subset(
  sub_seurat_no_cc,
  subset = orig.ident != "week12_036_extraadrenal" &
           orig.ident != "week9_031_paraganglia" &
           scDblFinder_keep == TRUE
)

```

  
Visualization to check populations and samples after filtering step. Moreover, checking the cell cycle status of population. Proliferating sympthatoblasts are present! 
  
```{r}
# UMAP colored by sample 
p1 <- DimPlot(sub_seurat_no_cc, 
              reduction = "umap", 
              group.by = "orig.ident", 
              pt.size = 0.3) + 
  ggtitle("UMAP: Samples") +
  theme(
    plot.title = element_text(size = 12)
  )

# UMAP colored by clusters 
p2 <- DimPlot(sub_seurat_no_cc, 
              reduction = "umap", 
              group.by = "cluster_label",
              label = TRUE, 
              repel = TRUE,
              pt.size = 0.3) + 
  ggtitle("UMAP: Clusters") +
  theme(
    plot.title = element_text(size = 12),
    legend.position = "none"
  )

# Combine plots side by side
p1 + p2 + 
  plot_layout(ncol = 2)
```


```{r}
DimPlot(sub_seurat_no_cc, reduction = "umap", group.by = 'Phase', label = FALSE, repel = TRUE) 

```

Saved as a backup step here.
```{r}
# save sub_seurat object and empty environment
#saveRDS(sub_seurat_no_cc, file = "/Users/adamadonyi/Downloads/GSE147821_combined_seurat_clusteres_annotated_sub_clean.rds")
```

Cleaned memory to be able to render the pdf

```{r}


rm(list = ls())

# Optionally, you can force garbage collection
gc()

sub_seurat_no_cc <- readRDS("/Users/adamadonyi/Downloads/GSE147821_combined_seurat_clusteres_annotated_sub_clean.rds")


```


# SECTION_6.A: trajectory analysis and heatmap.
6a. Perform the trajectory analysis between SCPs and chromaffin cells, SCPs and
sympathoblasts, and between chromaffin cells and sympathoblasts.
Plot the important gene changes along the trajectories on heatmaps.
You can use any trajectory analysis tool. Please explain your choice.

After re running the basic pipeline of scRNAseq 

```{r}
sub_seurat_no_cc <- NormalizeData(sub_seurat_no_cc)


# 2. Find variable features
sub_seurat_no_cc <- FindVariableFeatures(sub_seurat_no_cc)

# 3. Scale the data
sub_seurat_no_cc <- ScaleData(sub_seurat_no_cc)

# 4. Run PCA - as per paper, they used 30 components
sub_seurat_no_cc <- RunPCA(sub_seurat_no_cc)

# 5. Determine significant PCs --> use the 3
elb <- ElbowPlot(sub_seurat_no_cc)

# important step - tested without 
sub_seurat_no_cc <- FindNeighbors(sub_seurat_no_cc, dims = 1:10)
sub_seurat_no_cc <- FindClusters(sub_seurat_no_cc, resolution = 0.1)

# 6 Run UMAP
sub_seurat_no_cc <- RunUMAP(sub_seurat_no_cc, dims = 1:10)

```

According to the reference paper Conos integration was performed (with parameters k=30, k.self=5 and 30 PCA components) on the dataset. 
This analysis integrates scRNA-seq data across samples using Conos, followed by UMAP visualization and trajectory inference with Slingshot on all cell types. 
Differential gene expression along pseudotime is assessed using tradeSeq, identifying genes dynamically regulated in each lineage transition. 
A ternary analysis computes signature scores for each cell type and orders cells by transition gradients to visualize lineage-specific gene expression patterns. 
Heatmaps of marker and pseudotime-responsive genes shows molecular trajectories of neurodevelopmental transitions.





```{r}

# Extract cell type information for later use
cell_types <- sub_seurat_no_cc$cluster_label


# Split the Seurat object by sample/batch
seurat_list <- SplitObject(sub_seurat_no_cc, split.by = "gsm")  # Assuming 'gsm' is your batch identifier

# Convert each Seurat object to pagoda2
p2_list <- lapply(seurat_list, function(x) {
  # Extract count
  counts <- GetAssayData(x, slot = "counts")
  
  # Create pagoda2 object
  p2 <- pagoda2::Pagoda2$new(counts, log.scale = TRUE, n.cores = 4)
  
  # Process pagoda2 object
  p2$adjustVariance(plot = FALSE)
  p2$calculatePcaReduction(nPcs = 30, n.odgenes = 3000)
  
  return(p2)
})

# Create Conos object
con <- conos::Conos$new(p2_list)

# Run Conos integration with specified parameters from the paper
# k=30, k.self=5 and 30 PCA components
con$buildGraph(k = 30, k.self = 5, space = "PCA", ncomps = 30)

# Calculate UMAP embedding
con$embedGraph(method = "UMAP", n.neighbors = 200)  # Using 200 neighbors as specified

# Get the joint graph and UMAP embedding
joint_graph <- con$graph
umap_embedding <- con$embedding

# Create a new Seurat object with integrated data
# Extract the cell embeddings
cell_embeddings <- con$embedding

# Create a new dimensional reduction object for Seurat 
umap_reduction <- CreateDimReducObject(
  embeddings = cell_embeddings,
  key = "UMAP_",
  assay = DefaultAssay(sub_seurat_no_cc)
)

# Add the UMAP reduction to the Seurat object
sub_seurat_no_cc[["umap"]] <- umap_reduction

# Add back the original cell type 
sub_seurat_no_cc$cell_type <- cell_types

# Visualize the UMAP with cell types
p1 <- DimPlot(sub_seurat_no_cc, reduction = "umap", group.by = "cluster_label")
print(p1)

# Convert to SingleCellExperiment for slingshot
sce <- as.SingleCellExperiment(sub_seurat_no_cc)

# Use the UMAP embedding from Conos
reducedDims(sce)[["UMAP"]] <- cell_embeddings

# Add normalized counts if not already present<-make sure
if (!"logcounts" %in% assayNames(sce)) {
  sce <- logNormCounts(sce)
}

# Filter for cells
sce_subset <- sce[, sce$cluster_label %in% c("Sympathoblasts", "Chromaffin cells", "SCPs")]

#  markers from literature
marker_genes <- list(
  SCP = c("SOX10", "PLP1", "FOXD3", "FABP7", "S100B", "ERBB3", "NGFR", 
          "MBP", "MPZ", "COL2A1", "POSTN", "MOXD1", "GAS7"),
  
  Sympathoblasts = c("STMN2", "HAND2", "ELAVL4", "STMN4", "ISL1", 
                     "PRPH", "ELAVL2", "HMX1"),
  
  Chromaffin = c("GAS7", "CHGA", "CHGB", "INSM1", "PENK", "PNMT", "SLC35D3")
)


# gene expression changes along trajectory - main Function<----------------------------
analyze_trajectory <- function(sce_obj, trajectory_name) {
  # Get pseudotime values for the first lineage
  pseudotime <- sce_obj$slingPseudotime_1
  
  # Print vector length and Number of NA for pseudotime
  cat("Initial pseudotime vector length:", length(pseudotime), "\n")
  cat("Number of NA values in pseudotime:", sum(is.na(pseudotime)), "\n")
  
  # Filter cells with NA pseudotime (if any!!!)
  valid_cells <- !is.na(pseudotime)
  sce_filtered <- sce_obj[, valid_cells]
  pseudotime <- pseudotime[valid_cells]
  
  cat("After filtering, pseudotime vector length:", length(pseudotime), "\n")
  
  # Get normalized counts for valid cells 
  if ("logcounts" %in% assayNames(sce_filtered)) {
    counts <- assay(sce_filtered, "logcounts")
  } else {
    # If logcounts don't exist, normalize and log-transform <---- error suggested then worked
    sce_filtered <- logNormCounts(sce_filtered)
    counts <- assay(sce_filtered, "logcounts")
  }
  
  cat("Number of cells in count matrix:", ncol(counts), "\n")
  
  # Create cell weights vector with the same length as pseudotime
  cell_weights <- rep(1, length(pseudotime))
  
  cat("Length of cell_weights vector:", length(cell_weights), "\n")
  
    # has the same length as the number of cells
  pseudotime_matrix <- matrix(NA, nrow = ncol(counts), ncol = 1)
  pseudotime_matrix[, 1] <- pseudotime
  
  # TradeSeq expects a list of pseudotime values per lineag
  lineages <- list(Lineage1 = pseudotime)
  
  # Print information about the structure
  cat("Structure of lineages object:\n")
  cat("  - Number of lineages:", length(lineages), "\n")
  cat("  - Length of Lineage1:", length(lineages[[1]]), "\n")
  
  # Double-check dimensions!!!
  if (length(cell_weights) != length(lineages[[1]])) {
    stop("Dimension mismatch: cell_weights (", length(cell_weights), 
         ") and pseudotime (", length(lineages[[1]]), ") have different lengths")
  }
  
  if (ncol(counts) != length(lineages[[1]])) {
    stop("Dimension mismatch: counts (", ncol(counts), 
         " columns) and pseudotime (", length(lineages[[1]]), " values) have different lengths")
  }
  
  # Fit GAM
  set.seed(42) 
  
  # Try using a simpler approach first with fewer knots (only solution for previous 303 error)
  tryCatch({
    cat("Fitting GAM model with tradeSeq...\n")
    gam_fits <- fitGAM(counts = counts, 
                       pseudotime = lineages,
                       cellWeights = cell_weights,
                       verbose = TRUE,
                       parallel = FALSE, # Set to FALSE to avoid parallel issues
                       nknots = 3) # Start with fewer knots for simplicity
    
    cat("GAM model fitted successfully!\n")
    
    # Test for genes that change
    cat("Testing for genes associated with pseudotime...\n")
    assoc_res <- associationTest(gam_fits)
    
    # Get significantly changing genes 0.05for p
    sig_genes <- rownames(assoc_res)[assoc_res$pvalue < 0.05]
    cat("Number of significant genes (p < 0.05):", length(sig_genes), "\n")
    
    # If too many significant genes, take TOP1000
    if (length(sig_genes) > 100) {
      sig_genes <- rownames(assoc_res)[order(assoc_res$pvalue)][1:100]
      cat("Taking top 100 significant genes\n")
    }
    
    # If no significant genes, use top 50 genes via variance
    if (length(sig_genes) == 0) {
      cat("No significant genes found, using top 50 by variation\n")
      var_genes <- rowVars(counts)
      sig_genes <- names(sort(var_genes, decreasing = TRUE))[1:50]
    }
    
    # Predict expression along trajectory for visual
    cat("Predicting gene expression along trajectory...\n")
    pseudotime_seq <- seq(min(pseudotime, na.rm = TRUE),
                         max(pseudotime, na.rm = TRUE),
                         length.out = 100)
    
    # Use tradeSeq's predictSmooth
    pred_counts <- predictSmooth(gam_fits, 
                                gene = sig_genes,
                                pseudotimes = list(Lineage1 = pseudotime_seq),
                                nPoints = 100)
    
    # Scale each gene for better visual
    pred_counts_scaled <- t(scale(t(pred_counts)))
    
    # Create heatmap 
    gene_order <- order(apply(pred_counts, 1, which.max))
    
    # Create heatmap
    cat("Creating heatmap...\n")
    heatmap_title <- paste("Genes changing along", trajectory_name)
    heat_map <- pheatmap(pred_counts_scaled[gene_order, ], 
                       cluster_rows = FALSE, 
                       cluster_cols = FALSE,
                       show_colnames = FALSE,
                       main = heatmap_title, # needed to add!!!!
                       labels_row = rownames(pred_counts_scaled)[gene_order])
    
    # Return results
    cat("Analysis complete!\n")
    return(list(
      significant_genes = sig_genes,
      heatmap = heat_map,
      predicted_expression = pred_counts_scaled[gene_order, ]
    ))
  }, error = function(e) {
    cat("Error occurred during analysis:", conditionMessage(e), "\n")
    cat("Attempting alternative approach...\n")
    
    # Sort cells by pseudotime
    cell_order <- order(pseudotime)
    cell_position <- seq_along(pseudotime) / length(pseudotime)
    
    # Create a simple heatmap of the top variable genes
    var_genes <- rowVars(counts)
    top_var_genes <- names(sort(var_genes, decreasing = TRUE))[1:50]
    
    # Extract and scale expression
    top_gene_expr <- counts[top_var_genes, ]
    top_gene_expr_scaled <- t(scale(t(top_gene_expr)))
    
    # Create heatmap
    heatmap_title <- paste("Top variable genes along", trajectory_name, "(alternative method)")
    heat_map <- pheatmap(top_gene_expr_scaled[, cell_order], 
                       cluster_rows = TRUE, 
                       cluster_cols = FALSE,
                       show_colnames = FALSE,
                       main = heatmap_title)
    
    return(list(
      message = paste("Original analysis failed with error:", conditionMessage(e)),
      significant_genes = top_var_genes,
      heatmap = heat_map
    ))
  })
}

# Create a function to generate heatmap 
create_transition_heatmap <- function(sce_obj, marker_genes, transition_name) {
  # Get expression data
  if ("logcounts" %in% assayNames(sce_obj)) {
    expr <- assay(sce_obj, "logcounts")
  } else {
    sce_obj <- logNormCounts(sce_obj)
    expr <- assay(sce_obj, "logcounts")
  }
  
  # Flatten the marker genes list
  all_markers <- unlist(marker_genes)
  
  # Keep only genes in our dataset
  valid_markers <- intersect(all_markers, rownames(expr))
  
  # Get expression for marker genes
  marker_expr <- expr[valid_markers, ]
  
  # Scale expression
  marker_expr_scaled <- t(scale(t(marker_expr)))
  
  # Create heatmap
  heatmap_title <- paste("Marker genes in", transition_name, "transition")
  heat_map <- pheatmap(marker_expr_scaled,
                      cluster_rows = TRUE,
                      cluster_cols = FALSE,
                      show_colnames = FALSE,
                      main = heatmap_title)
  
  return(heat_map)
}

# 1. SCPs to Chromaffin cells
sce_scp_chrom <- sce_subset[, sce_subset$cluster_label %in% c("SCPs", "Chromaffin cells")]
sce_scp_chrom <- slingshot(sce_scp_chrom, clusterLabels = 'cluster_label', 
                          reducedDim = 'UMAP', start.clus = 'SCPs')

# 2. SCPs to Sympathoblasts
sce_scp_symp <- sce_subset[, sce_subset$cluster_label %in% c("SCPs", "Sympathoblasts")]
sce_scp_symp <- slingshot(sce_scp_symp, clusterLabels = 'cluster_label', 
                         reducedDim = 'UMAP', start.clus = 'SCPs')

# 3. Chromaffin cells to Sympathoblasts
sce_chrom_symp <- sce_subset[, sce_subset$cluster_label %in% c("Chromaffin cells", "Sympathoblasts")]
sce_chrom_symp <- slingshot(sce_chrom_symp, clusterLabels = 'cluster_label', 
                           reducedDim = 'UMAP', start.clus = 'Chromaffin cells')

# Trajectories
# For SCPs to Chromaffin cells
colors <- colorRampPalette(c("#E0E0E0", "#0000AA"))(100)
plot(reducedDims(sce_scp_chrom)[["UMAP"]][,c(1,2)], 
     col = colors[cut(sce_scp_chrom$slingPseudotime_1, breaks=100)],
     pch=16, cex=0.5, 
     main="Trajectory: SCPs to Chromaffin cells")
lines(SlingshotDataSet(sce_scp_chrom), lwd=2, col="black")

# For SCPs to Sympathoblasts
plot(reducedDims(sce_scp_symp)[["UMAP"]][,c(1,2)], 
     col = colors[cut(sce_scp_symp$slingPseudotime_1, breaks=100)],
     pch=16, cex=0.5, 
     main="Trajectory: SCPs to Sympathoblasts")
lines(SlingshotDataSet(sce_scp_symp), lwd=2, col="black")

# For Chromaffin cells to Sympathoblasts
plot(reducedDims(sce_chrom_symp)[["UMAP"]][,c(1,2)], 
     col = colors[cut(sce_chrom_symp$slingPseudotime_1, breaks=100)],
     pch=16, cex=0.5, 
     main="Trajectory: Chromaffin cells to Sympathoblasts")
lines(SlingshotDataSet(sce_chrom_symp), lwd=2, col="black")

# analysis for each transition
# 1. SCPs to Chromaffin cells
scp_chrom_results <- analyze_trajectory(sce_scp_chrom, "SCPs to Chromaffin cells")

# 2. SCPs to Sympathoblasts
scp_symp_results <- analyze_trajectory(sce_scp_symp, "SCPs to Sympathoblasts")

# 3. Chromaffin cells to Sympathoblasts
chrom_symp_results <- analyze_trajectory(sce_chrom_symp, "Chromaffin cells to Sympathoblasts")

# Perform ternary analysis as described in the paper
perform_ternary_analysis <- function(sce_obj, marker_genes) {
  # Get expression data
  if ("logcounts" %in% assayNames(sce_obj)) {
    expr <- assay(sce_obj, "logcounts")
  } else {
    sce_obj <- logNormCounts(sce_obj)
    expr <- assay(sce_obj, "logcounts")
  }
  
  # Calculate signature scores for each cell type
  calc_signature_score <- function(gene_list) {
    # Keep only genes in the dataset
    valid_genes <- intersect(gene_list, rownames(expr))
    if (length(valid_genes) == 0) {
      return(rep(0, ncol(expr)))
    }
    # Sum exp values
    colSums(expr[valid_genes, , drop = FALSE])
  }
  
  # Calculate scores
  scp_score <- calc_signature_score(marker_genes$SCP)
  symp_score <- calc_signature_score(marker_genes$Sympathoblasts)
  chrom_score <- calc_signature_score(marker_genes$Chromaffin)
  
  # Scores to metadata
  sce_obj$scp_score <- scp_score
  sce_obj$symp_score <- symp_score
  sce_obj$chrom_score <- chrom_score
  
  # Calculate ratios for ordering (following paper methodology)
  total_score <- scp_score + symp_score + chrom_score
  sce_obj$chrom_ratio <- chrom_score / total_score
  
  # For SCP to Sympathoblast transition (as in Fig 2h)
  # Select cells where chromaffin signature is not dominant
  # Handle NAs by replacing them with FALSE!!!!
  chrom_filter <- sce_obj$chrom_ratio <= 0.7
  chrom_filter[is.na(chrom_filter)] <- FALSE
  
  scp_to_symp <- sce_obj[, chrom_filter]
  
  # Calculate ratio only for filtered cells
  symp_to_sum <- scp_to_symp$symp_score / (scp_to_symp$symp_score + scp_to_symp$scp_score)
  # Handle division by zero or other NAs
  symp_to_sum[is.na(symp_to_sum)] <- 0
  scp_to_symp$symp_to_sum_ratio <- symp_to_sum
  
  # Order cells by this ratio
  cell_order <- order(scp_to_symp$symp_to_sum_ratio)
  
  # Similar analysis for SCP to Chromaffin transition
  # Handle NAs in the filter
  symp_filter <- sce_obj$symp_score / total_score <= 0.7
  symp_filter[is.na(symp_filter)] <- FALSE
  
  scp_to_chrom <- sce_obj[, symp_filter]
  
  # Calculate ratio only for filtered cells
  chrom_to_sum <- scp_to_chrom$chrom_score / (scp_to_chrom$chrom_score + scp_to_chrom$scp_score)
  # Handle division by zero or other NAs
  chrom_to_sum[is.na(chrom_to_sum)] <- 0
  scp_to_chrom$chrom_to_sum_ratio <- chrom_to_sum
  
  chrom_cell_order <- order(scp_to_chrom$chrom_to_sum_ratio)
  
  # And for Sympathoblast to Chromaffin transition (as in Fig 3b)
  # Handle NAs in the filter
  scp_filter <- sce_obj$scp_score / total_score <= 0.7
  scp_filter[is.na(scp_filter)] <- FALSE
  
  symp_to_chrom <- sce_obj[, scp_filter]
  
  # Calculate ratio only for filtered cells
  chrom_to_symp <- symp_to_chrom$chrom_score / (symp_to_chrom$chrom_score + symp_to_chrom$symp_score)
  # Handle division by zero or other NAs
  chrom_to_symp[is.na(chrom_to_symp)] <- 0
  symp_to_chrom$chrom_to_symp_ratio <- chrom_to_symp
  
  symp_chrom_cell_order <- order(symp_to_chrom$chrom_to_symp_ratio)
  
  # Return the ordered cells for each transition
  return(list(
    scp_to_symp = scp_to_symp[, cell_order],
    scp_to_chrom = scp_to_chrom[, chrom_cell_order],
    symp_to_chrom = symp_to_chrom[, symp_chrom_cell_order]
  ))
}

# Execute the ternary analysis (this was missing in the original code)
ternary_results <- perform_ternary_analysis(sce_subset, marker_genes)

#  create the heatmaps
scp_to_symp_heatmap <- create_transition_heatmap(ternary_results$scp_to_symp, marker_genes, "SCP to Sympathoblast")
scp_to_chrom_heatmap <- create_transition_heatmap(ternary_results$scp_to_chrom, marker_genes, "SCP to Chromaffin")
symp_to_chrom_heatmap <- create_transition_heatmap(ternary_results$symp_to_chrom, marker_genes, "Sympathoblast to Chromaffin")

# Extract top genes from each transition
top_genes_per_transition <- list(
  SCPs_to_Chromaffin = head(scp_chrom_results$significant_genes, 10),
  SCPs_to_Sympathoblasts = head(scp_symp_results$significant_genes, 10),
  Chromaffin_to_Sympathoblasts = head(chrom_symp_results$significant_genes, 10)
)

# Create a summary table
cat("Top genes changing in each transition:\n")
for (transition in names(top_genes_per_transition)) {
  cat("\n", transition, ":\n")
  cat(paste(top_genes_per_transition[[transition]], collapse = ", "), "\n")
}


```

The umap shows only the 3 cell types we were interested. 

Trajectory graphs are color coded the pseudotime and the Singleshot provided line represent the most likely path cells take during differentiation.


SCP to Chromaffin cells: gradually acquire chromaffin-like-cell identity through a relatively direct developmental trajectory.

SCP to Sympathoblasts:The curve has a distinct shape compared to the chromaffin trajectory, indicating these are separate developmental programs with different gene expression patterns. The pronounced curve suggests intermediate cell states may play important roles in this transition

Chromaffin cells to Sympathoblasts: We can assume based on the plot there is a relationship between chromaffin cells and sympathoblasts. The trajectory suggests these cell types might share developmental history but diverge along different paths.

Hatmap of marker genes shows the key  signature genes to investigate potential transitions over pseudotime.

SCP to Sympathoblast Transition: Expression of SCPs markers such as SOX10, FOXD3, POSTN  decrease moving rightward.Genes like STMN2, HAND2, ELAVL4, and HMX1 show increased expression toward the right side demonstrating the acquired Sympathoblast properties.  This indicates a clear developmental trajectory from SCPs (left) to Sympathoblasts (right) with a gradual loss of SCP identity genes and gain of sympathoblast markers.

SCP to Chromaffin Transition: Similar to first heatmap, SCP markers like SOX10, FOXD3, POSTN show high expression on the left and decrease toward the right.Genes like CHGA, CHGB, PNMT, PENK show low expression in SCPs but increased expression on the right, representing chromaffin differentiation. This transition demonstrates a different cellular fate from the same SCP progenitors, with the acquisition of chromaffin-specific markers.


Sympathoblast to Chromaffin Transition:
HAND2, HMX1, ISL1 are highly expressed on the left (sympathoblasts). Chromaffin markers such as CHGA, CHGB, PNMT, PENK increase toward the right side (middle section of the heatmap) This suggests potential plasticity between sympathoblast and chromaffin fates, which is interesting from a developmental perspective.


